Towards practical dynamic induction control of wind farms: analysis of optimally controlled wind-farm boundary layers and sinusoidal induction control of first-row turbines

### 2018-05-23 revision/reply_to_revs_2.tex
Feedback Johan part 1

\usepackage{graphicx}

@@

\textbf{Response:} The reason for this different behavior of the downstream vs upstream turbines is twofold:
\begin{enumerate}
	\item The Extremum Seeking Control (ESC) approach by Ciri \emph{et al.} differs fundamentally from our approach. In our approach, we simultaneously optimize all control variables over a given time horizon, which allows for full dynamic cooperation and interaction between all turbines. In the ESC, controls are optimized one by one in the sense that ESC is initiated on the first row turbine and will be activated on the subsequent row only after its upstream neighbor has converged to a steady state solution. In this way, dynamic control and interaction between controls in different turbines are discarded a priori. Furthermore, although this is not an inherent limitation of the ESC approach, Ciri \emph{et al.} optimize for individual turbine power instead of total wind-farm power (see Eq. 10 $P_i$ is used instead of $\sum_i P_i$). This is explicitly mentioned in the title of their follow up paper (Ciri \emph{et al.} Large-eddy simulations with extremum-seeking control for individual wind turbine power optimization)

	\item In our modeling approach, we make abstraction of detailed turbine dynamics in the sense that we directly control the thrust coefficient $C_T'$ instead of the generator torque or tip speed ratio. In this sense, the optimal thrust coefficient value of a turbine with no downstream neighbors can be analytically found to be $C_T' = 2$. In Ciri \emph{et al.}, the authors indicate that the reason why the TSR in the downstream rows (around 6.5) does not correspond to the individual optimal TSR (7.5) of individual turbines is that the TSR depends on a reference velocity and (direct quotation) ``As a consequence, the TSR is not a meaningful indicated of a waked turbine operating condition".

\end{enumerate}

Summarizing, the inverse behavior of the control dynamics compared to Ciri \emph{et al.} can be attributed to the difference in approach to wind-farm control (dynamic full-farm optimization vs steady sequential and individual optimization) as well as the implementation of the turbine control ($C_T'$ vs $\lambda$). In our opinion, the current study (full-farm dynamic wind-farm optimization) and that of Ciri \emph{et al.} (providing a practical ESC method for optimizing individual turbines) cannot be directly compared.

However, we understand the contrast in the results and therefore added a small comment to the revised manuscript  (l. ??, p. ??):

``
... is not monotonous. \textbf{Note that these observations are in contrast to the study of Ciri et al. 2017, where extremum seeking control is applied to optimize individual turbines in an array of three turbines. They found that set points in the first-row turbines remain untouched from their locally optimal values whereas downstream turbines are re-optimized for wake conditions. This contrast can be attributed to our dynamic and simultaneous wind-farm control approach, which differs from their sequential individual set-point control.  Furthermore, self-optimization in the latter study is achieved by recalibrating the torque gain. The baseline controller in our current actuator disk model makes abstraction of pitch and generator torque control and keeps $C_T'$ at its theoretical optimum instead.}
''

WM - WEET NIET OF WE DIT WEL ECHT MOETEN DOEN, MISSCHIEN GEWOON REVIEWER OVERTUIGEN DAT WE DIT NIET INCLUDEN? TE BESPREKEN

@@

Current turbines are designed to approach maximum $C_T'$ values around 2, corresponding to the Betz limit. Although this maximum value can be increased by increasing the operational tip speed ratio (TSR), aiming to do so by adapting the pitch angle would inevitably lead to severe efficiency losses due to stall on the turbine blades. Therefore, we provide an example of how an alternative turbine design (i.e. with an increased chord length and operational TSR) could attain a maximum $C_T'$ of 3.5. Given such a turbine design, achieving thrust ratings $C_T' < 3.5$ is straightforward by pitching blades towards the feather position.

@@

``
... be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than 
the original design value (see Goit and Meyers, 2015, Appendix A). \textbf{Furthermore, given this redesign, dynamic reductions from this maximum value could be realized through pitch control. More specifically, based on speculated specifications of similar turbines in practice, the NREL 5 MW is designed with a pitch rate of $8^{\circ}/$s. This implies that the rotor can go from power maximization ($0^\circ$ pitch) to total curtailment in fully feathered blade position ($90^\circ$ pitch) in 11.25 seconds (Jonkman et al. 2009).}
''

@@

\textbf{Response: } Indeed, the high inertia of the rotor could be a compromise to achieving the desired sinusoidal thrust coefficient behavior. However, given the proposed redesign of the NREL 5MW rotor, dynamic reductions from the maximum $C_T' = 3.5$ could be achieved rapidly by using the fast pitch actuators with which modern turbines are equipped. We have added this comment to the discussion as follows:

``
... be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than 
the original design value (see Goit and Meyers, 2015, Appendix A). \textbf{Furthermore, given this redesign, dynamic reductions from this maximum value could be realized through pitch control. More specifically, based on speculated specifications of similar turbines in practice, the NREL 5 MW is designed with a pitch rate of $8^{\circ}/$s. This implies that the rotor can go from power maximization ($0^\circ$ pitch) to total curtailment in fully feathered blade position ($90^\circ$ pitch) in 11.25 seconds (Jonkman et al. 2009).}
''

@@

Moreover, it can be seen that the full optimal control (C3t5 in Figure 20) however is able to contain the high momentum zone by continuously improving the mixing process with the background boundary layer. The mechanisms thereof however remain elusive to date, and are subject of further investigation.

>>>>

\usepackage{graphicx}
\usepackage{color}

@@

\textbf{Response:} This is an interesting comment. It is correct that this contrast results from the way turbines and their controls are modeled in the current study and that of Ciri et al. The reference disk-based thrust coefficient $C_T' = 2$ used here actually implies intrinsic self-optimization of blade pitch and generator torque to local flow conditions, even though these actions are not resolved by our current actuator disk formulation which directly controls $C_T'$. In case these degrees of freedom would be resolved, for instance using an actuator line model as in Ciri et al., the ESC from the latter study could e.g. be used to optimize torque controller gain in order to achieve an effective $C_T' \approx 2$.

We have included the fact that the reference case already implies self-optimization in the revised manuscript as follows (p 4, l. 21):

``
A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ was defined as a reference case. \textbf{Note that this would correspond to a farm with ideal turbines for which generator torque is being controlled dynamically to track the maximum power point at the Betz limit perfectly.} Several different optimal control cases were defined, ...
''

However, even given the discussion above, it is not unimaginable that a turbine could temporarily outperform this greedy control setting, e.g. by timely reacting to incoming flow accelerations. This is what we would refer to as self-optimization. {\color{blue} We have explicitly defined this in the text as follows (p ??):

``
...This suggests that self-optimization \textbf{(i.e. temporarily outperforming conventional greedy control in response to incoming flow events)} is very limited: the optimal controls for a given turbine are designed to create ...
'' }

(IS DIT OKE?? Kan het deel in blauw misschien beter gewoon weglaten?)

@@

Current turbines are designed to approach maximum $C_T'$ values around 2, corresponding to the Betz limit. Aiming to increase thrust by adapting the pitch angle would inevitably lead to severe efficiency losses due to stall on the turbine blades. Therefore, we provide an example of how an alternative turbine design (i.e. with an increased chord length and operational TSR) could attain a maximum $C_T'$ of 3.5. Given such a turbine design, achieving thrust ratings $C_T' < 3.5$ is straightforward by pitching blades towards the feather position.

@@

``
The resulting maximum thrust coefficient of 3.5 can, e.g., be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than 
the original design value (see Goit and Meyers, 2015, Appendix A). \textbf{Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation in the order of $10^\circ/$s is possible (see, e.g., Jonkman et al. 2009).} 
''

Note that this control strategy is not unique and is solely given as an indication for the technical feasibility of tracking the proposed $C_T'$ waveform. In practice, we expect that this can also be achieved through a combination of generator torque and blade pitch control. This is subject of ongoing investigation.

@@

\textbf{Response: } We thank the reviewer for this suggestion. Dynamic reductions from the maximum $C_T' = 3.5$ could, for instance, be achieved by using the fast pitch actuators with which modern turbines are equipped. We have added this comment to the discussion as follows (p ??):

``
The resulting maximum thrust coefficient of 3.5 can, e.g., be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than 
the original design value (see Goit and Meyers, 2015, Appendix A). \textbf{Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation in the order of $10^\circ/$s is possible (see, e.g., Jonkman et al. 2009).} 
''

Note that this control strategy is not unique and is solely given as an indication for the technical feasibility of tracking the proposed $C_T'$ waveform. In practice, we expect that this can also be achieved through a combination of generator torque and blade pitch control. This is subject of ongoing investigation.

@@

Moreover, it can be seen that the full optimal control (C3t5 in Figure 20) is able to contain the high momentum zone by continuously improving the mixing process with the background boundary layer. The mechanisms thereof however remain elusive to date, and are subject of further investigation.

### 2018-05-23 revision/revised_WM02.tex
Feedback Johan part 1

\newcommand{\revision}[1]{{\color{red} #1}}

@@

A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ (corresponding to the Betz limit in momentum theory) was defined as a reference case. Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq. \ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or~3 \revision{, with thrust forces that can respectively only be reduced (underinductive), or also increased (overinductive) compared to the Betz optimum at $C_T' = 2$} (see Eq. \ref{eq:boxct_constraint}). Cases are denoted as C<X>t<Y>, where X and Y represent $C_{T,\text{max}}'$ and $\tau$ respectively, e.g. C3t30  for the case with $C_{T,\text{max}}' = 3$ and $\tau = 30$ s. \revision{The choice of (and sensitivity to) setup parameters is further elaborated in MM17.}

@@

Figure~\ref{fig:multirow}b shows results from simulations in which the controls are applied \emph{for all rows up to a certain row}, i.e. R1--R3 indicates the application of optimized controls generated by case C3t5 to rows 1, 2, and 3. An interesting observation from this figure is that , for any row $i$ except the last one, the power potential as observed in case C3t5 is almost fully recovered by only applying the optimal controls up to row $i-1$. This suggests that self-optimization is very limited: the optimal controls for a given turbine are designed to create favorable flow conditions in the downstream rows instead of increasing local power. Furthermore, although the discussion in the previous paragraph has shown that downstream controls are optimized with the upstream actions in mind, the converse is not true: upstream control actions do not require a specific downstream response in order to increase power in that downstream row.

@@

Figure~\ref{fig:sinus_baseline} illustrates the power extraction for all cases considered. Figure~\ref{fig:sinus_baseline}a illustrates the relative power gains over the reference case. From the figure it can be seen that there is a well-defined range of values for $A$ and $St$ for which wind-farm power can be increased substantially through upstream sinusoidal thrust variations, with a maximal power increase of $\approx$ 5\% at $(St^\star, A^\star) = (0.25, 1.5)$. For instance, a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$. The resulting maximum thrust coefficient of 3.5 can, e.g., be \revision{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in chord length at a tip speed ratio that is 25\% higher than the greedy tip speed ratio (see \citealp{goit2015optimal}, Appendix A). \revision{Furthermore, given this redesign, dynamic reductions from this maximum value could be realized through pitch control. More specifically, based on speculated specifications of similar turbines in practice, the NREL 5 MW is designed with a pitch rate of $8^{\circ}/$s. This implies that the rotor can go from power maximization ($0^\circ$ pitch) to total curtailment in fully feathered blade position ($90^\circ$ pitch) in 11.25 seconds \cite{jonkman2009definition}.} Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17. The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%. In contrast, optimal control case C3t5, in which all rows are active, also increases power in rows 3 and 4, and reduces first-row power by only 1\%. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figure14}
	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_baseline} }
\end{figure}

>>>>

\newcommand{\revision}[1]{{\color{blue} #1}}

@@

A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ was defined as a reference case. \revision{Note that this would correspond to a farm with ideal turbines for which generator torque is being controlled dynamically to track the maximum power point at the Betz limit perfectly.} Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq. \ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or~3 \revision{, with thrust forces that can respectively only be reduced (underinductive), or also increased (overinductive) compared to the Betz optimum at $C_T' = 2$} (see Eq. \ref{eq:boxct_constraint}). Cases are denoted as C<X>t<Y>, where X and Y represent $C_{T,\text{max}}'$ and $\tau$ respectively, e.g. C3t30  for the case with $C_{T,\text{max}}' = 3$ and $\tau = 30$ s. \revision{The choice of (and sensitivity to) setup parameters is further elaborated in MM17.}

@@

Figure~\ref{fig:multirow}b shows results from simulations in which the controls are applied \emph{for all rows up to a certain row}, i.e. R1--R3 indicates the application of optimized controls generated by case C3t5 to rows 1, 2, and 3. An interesting observation from this figure is that , for any row $i$ except the last one, the power potential as observed in case C3t5 is almost fully recovered by only applying the optimal controls up to row $i-1$. This suggests that self-optimization \revision{(i.e. temporarily outperforming conventional greedy control in response to incoming flow events)} is very limited: the optimal controls for a given turbine are designed to create favorable flow conditions in the downstream rows instead of increasing local power. Furthermore, although the discussion in the previous paragraph has shown that downstream controls are optimized with the upstream actions in mind, the converse is not true: upstream control actions do not require a specific downstream response in order to increase power in that downstream row.

@@

Figure~\ref{fig:sinus_baseline} illustrates the power extraction for all cases considered. Figure~\ref{fig:sinus_baseline}a illustrates the relative power gains over the reference case. From the figure it can be seen that there is a well-defined range of values for $A$ and $St$ for which wind-farm power can be increased substantially through upstream sinusoidal thrust variations, with a maximal power increase of $\approx$ 5\% at $(St^\star, A^\star) = (0.25, 1.5)$. For instance, a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$. The resulting maximum thrust coefficient of 3.5 can, e.g., be \revision{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in chord length at a tip speed ratio that is 25\% higher than the greedy tip speed ratio (see \citealp{goit2015optimal}, Appendix A). \revision{Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation rates in the order of $10^\circ/$s are possible (see, e.g., \citealp{jonkman2009definition}).} Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17. The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%. In contrast, optimal control case C3t5, in which all rows are active, also increases power in rows 3 and 4, and reduces first-row power by only 1\%. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figure14}
	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_baseline} }
\end{figure}
